---
title: "DATA*6500 Course Project Report"
author: "Evan Switzer, Namrata Roy"
date: "2024-07-11"
output:
  html_document: default
  word_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, echo=FALSE, results='hide', message=FALSE, warning=FALSE}
library(dplyr)
library(tmap)
library(osmdata)
library(sf)
library(rosm)
library(tmaptools)
library(ggplot2)
library(tinytex)
library(tidyr)
library(leaflet)
library(spatstat)
library(raster)
library(spdep)
library(lubridate)
library(INLA)
library(caret)
library(randomForest)
library(xgboost)
```

# Data Sources

## Neighbourhood_Crime_Rates_Open_Data
#### https://data.torontopolice.on.ca/datasets/TorontoPS::neighbourhood-crime-rates-open-data/about



```{r, echo=FALSE, results='hide', message=FALSE, warning=FALSE}
#input data
NeighbourhoodCrimeRates = read.csv("Neighbourhood_Crime_Rates_Open_Data.csv")

MajorCrimeIndicators = read.csv("Major_Crime_Indicators_Open_Data.csv")

BicycleThefts = read.csv("Bicycle_Thefts_Open_Data.csv")
```

```{r}
#for BicycleThefts plot each occurance of BIKE_COST
ggplot(BicycleThefts, aes(x = BIKE_COST)) +
  geom_histogram(binwidth = 100,color = "black", alpha = 0.7) +
  labs(title = "Distribution of Bike Costs",
       x = "Bike Cost ($)",
       y = "Count") +
  theme_minimal()

print(summary(BicycleThefts$BIKE_COST))
#replace na with mean
BicycleThefts$BIKE_COST[is.na(BicycleThefts$BIKE_COST)] <- mean(BicycleThefts$BIKE_COST, na.rm = TRUE)

BicycleThefts <- BicycleThefts |>
  filter_all(all_vars(. != "NSA"))

#print number of 0s, 0s are left because some bikes are actually worth very little (like my bike)
print(sum(BicycleThefts$BIKE_COST == 0))
```

```{r}
#barplot of BIKE_TYPE
ggplot(BicycleThefts, aes(x = BIKE_TYPE)) +
  geom_bar() +
  labs(title = "Bike Types",
       x = "Bike Type",
       y = "Count") +
  theme_minimal()

#barplot of BIKE_COLOUR
ggplot(BicycleThefts, aes(x = BIKE_COLOUR)) +
  geom_bar() +
  labs(title = "Bike Colours",
       x = "Bike Colour",
       y = "Count") +
  theme_minimal()

#barplot of BIKE_MAKE
ggplot(BicycleThefts, aes(x = BIKE_MAKE)) +
  geom_bar() +
  labs(title = "Bike Makes",
       x = "Bike Make",
       y = "Count") +
  theme_minimal()

#barplot of BIKE_MODEL
ggplot(BicycleThefts, aes(x = BIKE_MODEL)) +
  geom_bar() +
  labs(title = "Bike Models",
       x = "Bike Model",
       y = "Count") +
  theme_minimal()

#barplot of BIKE_SPEED
ggplot(BicycleThefts, aes(x = BIKE_SPEED)) +
  geom_bar() +
  labs(title = "Bike Speeds",
       x = "Bike Speed",
       y = "Count") +
  theme_minimal()

#barplot of PREMISES_TYPE
ggplot(BicycleThefts, aes(x = PREMISES_TYPE)) +
  geom_bar() +
  labs(title = "Premises Types",
       x = "Premises Type",
       y = "Count") +
  theme_minimal()

#barplot of LOCATION_TYPE
ggplot(BicycleThefts, aes(x = LOCATION_TYPE)) +
  geom_bar() +
  labs(title = "Location Types",
       x = "Location Type",
       y = "Count") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

#barplot of primary_offence
ggplot(BicycleThefts, aes(x = PRIMARY_OFFENCE)) +
  geom_bar() +
  labs(title = "Primary Offences",
       x = "Primary Offence",
       y = "Count") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```

```{r}
#for location types plot top 10
BicycleThefts %>%
  count(LOCATION_TYPE) %>%
  arrange(desc(n)) %>%
  head(10) %>%
  ggplot(aes(x = reorder(LOCATION_TYPE, n), y = n)) +
  geom_bar(stat = "identity") +
  labs(title = "Top 10 Location Types for Bike Thefts",
       x = "Location Type",
       y = "Count") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1, size = 8))

#for primary offence plot top 5
BicycleThefts %>%
  count(PRIMARY_OFFENCE) %>%
  arrange(desc(n)) %>%
  head(5) %>%
  ggplot(aes(x = reorder(PRIMARY_OFFENCE, n), y = n)) +
  geom_bar(stat = "identity") +
  labs(title = "Top 5 Primary Offences for Bike Thefts",
       x = "Primary Offence",
       y = "Count") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1, size = 8))


```


```{r}

# Parse OCC_DATE to POSIXct
BicycleThefts <- BicycleThefts %>%
  mutate(OCC_DATE = ymd_hms(OCC_DATE))

#for bike thefts remove any with OCC_YEAR less than 2014, greater than or equal to 2024, or NA
BicycleThefts <- BicycleThefts %>%
  filter(!is.na(OCC_YEAR), OCC_YEAR >= 2014, OCC_YEAR < 2024)
```

```{r}


#create days from start of 2014 for each theft
BicycleThefts$OCC_DATE_FROM2014 <- as.numeric(difftime(BicycleThefts$OCC_DATE, as.Date("2014-01-01"), units = "days"))

#summarize thefts per day
BicycleTheftsPerDay <- BicycleThefts %>%
  group_by(OCC_DATE_FROM2014) %>%
  summarize(thefts = n()) 

#summarize thefts per month per year
BicycleTheftsPerMonth <- BicycleThefts %>%
  mutate(OCC_MONTH = month(OCC_DATE)) %>%
  group_by(OCC_MONTH, OCC_YEAR) %>%
  summarize(thefts = n()) 

BicycleTheftsPerMonth <- BicycleTheftsPerMonth %>%
  mutate(YearMonth = make_date(OCC_YEAR, OCC_MONTH, 1))

BicycleTheftsPerMonth$time_id <- as.integer(as.factor(BicycleTheftsPerMonth$YearMonth))
BicycleTheftsPerMonth$time_id2 <- BicycleTheftsPerMonth$time_id

```


```{r, echo=FALSE, results='hide', message=FALSE, warning=FALSE}
#get background map of Toronto using openstreetmap
toronto_bbox = c(left = -79.6392, 
                 bottom = 43.403221,
                 right = -79.115952,
                 top = 43.855457)

toronto_border = opq(toronto_bbox, timeout = 600) |> 
  add_osm_feature(key = "boundary", value = "administrative") |> 
  add_osm_feature(key = 'admin_level', value = '6') |> 
  osmdata_sf() |> 
  (\(x) x$osm_multipolygons)() |>
  filter(name == "Toronto") |> 
  dplyr::select()

toronto_ward = opq(toronto_bbox, timeout = 60) |> 
  add_osm_feature(key = 'admin_level', value = '9') |> 
  osmdata_sf() |> 
  (\(x) x$osm_multipolygons)()

subwayStops = opq(toronto_bbox) |> 
  add_osm_feature(key = 'railway', value = 'subway_entrance') |> 
  osmdata_sf() |> 
  (\(x) x$osm_points)() |> 
  dplyr::select(name)

#for subway stops add the subway line
subwayLines = opq(toronto_bbox) |> 
  add_osm_feature(key = 'railway', value = 'subway') |> 
  osmdata_sf() |> 
  (\(x) x$osm_lines)() |> 
  dplyr::select(name)

toronto_subwayLines = subwayLines[toronto_border,, op=st_within]

#for each subwayStop find the subway line
subwayStops <- subwayStops |>
  mutate(subwayLineIndex = st_nearest_feature(subwayStops, toronto_subwayLines),
         subwayLine = toronto_subwayLines$name[subwayLineIndex])

toronto_subwayStops = subwayStops[toronto_border,, op=st_within] 
```


```{r}
#get bus stops from openstreetmap
busStops = opq(toronto_bbox) |> 
  add_osm_feature(key = 'highway', value = 'bus_stop') |> 
  osmdata_sf() |> 
  (\(x) x$osm_points)() |> 
  dplyr::select(name)

toronto_busStops = busStops[toronto_border,, op=st_within]

#get bike parking from openstreetmap
bikeParking = opq(toronto_bbox) |> 
  add_osm_feature(key = 'amenity', value = 'bicycle_parking') |> 
  osmdata_sf() |> 
  (\(x) x$osm_points)() |> 
  dplyr::select(name)

toronto_bikeParking = bikeParking[toronto_border,, op=st_within]

```

```{r}
#read in 3 xlsx files for population data (25-WardNames-Numbers, 2023-WardProfiles-GeographicAreas, population)
wardProfiles = readxl::read_xlsx("2023-WardProfiles-GeographicAreas.xlsx")
population = readxl::read_xlsx("population.xlsx")

wardProfiles <- wardProfiles |> 
  mutate(Ward = as.character(Ward))

#change column name of Area (sq km) to Area_sq_km
wardProfiles <- wardProfiles |> 
  rename(Area_sq_km = `Area (sq km)`)
```

```{r}
#get toronto wards from open street map
toronto_ward = opq(toronto_bbox, timeout = 60) |> 
  add_osm_feature(key = 'admin_level', value = '9') |> 
  osmdata_sf() |> 
  (\(x) x$osm_multipolygons)()

#combine ward names, profiles, population and toronto wards
toronto_ward_info = toronto_ward |>
  left_join(wardProfiles, by = c("ref" = "Ward"))

total_population <- t(population)
total_population <- total_population[-1, ]  # Remove the first row (the header)
total_population <- as.data.frame(total_population)  # Convert matrix to data frame

# Keep only the first column
total_population <- total_population[, 1, drop = FALSE]

#remove first row again
Toronto_pop <- total_population[1, ]
total_population <- total_population[-1, , drop = FALSE]

# Add Ward column with row numbers as Ward numbers
total_population$Ward <- 1:nrow(total_population)

# Rename the columns
colnames(total_population) <- c("Total_Population", "Ward")

#change total_population Ward into character
total_population$Ward <- as.character(total_population$Ward)

#merge with toronto ward info
toronto_ward_info = toronto_ward_info |>
  left_join(total_population, by = c("ref" = "Ward"))

# Convert the columns to numeric if they are not already
toronto_ward_info$Total_Population <- as.numeric(toronto_ward_info$Total_Population)
toronto_ward_info$Area_sq_km <- as.numeric(toronto_ward_info$Area_sq_km)

#calculate population density
toronto_ward_info$Population_Density = toronto_ward_info$Total_Population / toronto_ward_info$Area_sq_km

#remove columns: admin_level, boundary, political_division, type, wikidata, wikipedia, place
toronto_ward_info <- toronto_ward_info |>
  dplyr::select(-c(admin_level, boundary, political_division, type, wikidata, wikipedia, place))

#change ref column name to ward_id
toronto_ward_info <- toronto_ward_info |>
  rename(ward_id = ref)


```



```{r}
#plot toronto wards with population density
tm_shape(toronto_ward_info) +
  tm_polygons(col = "Population_Density", style = "quantile", palette = "Blues", title = "Population Density (people/sq km)") +
  tm_layout(legend.position = c("left", "bottom"))

#plot toronto wards with names on each of them
tm_shape(toronto_ward_info) +
  tm_polygons(col = "Population_Density", style = "quantile", palette = "Blues", title = "Population Density (people/sq km)") +
  tm_text("ward_id", size = 1) +
  tm_layout(legend.position = c("left", "bottom"))


# Convert to sf object
BicycleThefts_sf <- st_as_sf(BicycleThefts, 
                             coords = c("LONG_WGS84", "LAT_WGS84"), 
                             crs = 4326)  # Adjust the CRS as needed

tm_shape(toronto_ward_info) +
  tm_polygons(col = "Population_Density", style = "quantile", palette = "Blues", title = "Population Density (people/sq km)") +
  tm_shape(BicycleThefts_sf) +
  tm_dots(col = "red", size = 0.1, alpha = 0.2) +
  tm_layout(legend.position = c("left", "bottom"))

```


```{r}
#for each BicycleTheft, find the neighbourhood that each observation is in and add it to the data
BicycleThefts_with_wards <- st_join(BicycleThefts_sf, toronto_ward_info)
```

```{r}
#aggregate counts of theft by ward
BicycleThefts_by_ward <- BicycleThefts_with_wards %>%
  group_by(ward_id) %>%
  summarize(thefts = n())

#join with toronto ward info
toronto_ward_info <- st_join(toronto_ward_info, BicycleThefts_by_ward, left = TRUE)
#drop ward_id.x
toronto_ward_info <- toronto_ward_info |>
  dplyr::select(-c(ward_id.x))
#rename ward_id.y to ward_id
toronto_ward_info <- toronto_ward_info |>
  rename(ward_id = ward_id.y)

#calculate thefts per person
toronto_ward_info$Thefts_per_person = toronto_ward_info$thefts / toronto_ward_info$Total_Population

#calculate thefts per sq km
toronto_ward_info$Thefts_per_sq_km = toronto_ward_info$thefts / toronto_ward_info$Area_sq_km

#calculate thefts per pop density
toronto_ward_info$Thefts_per_pop_density = toronto_ward_info$thefts / toronto_ward_info$Population_Density



#plot thefts by ward
tm_shape(toronto_ward_info) +
  tm_polygons(col = "thefts", style = "cont", palette = "Blues", title = "Thefts") +
  tm_layout(legend.position = c("left", "bottom"))

#plot thefts by ward
tm_shape(toronto_ward_info) +
  tm_polygons(col = "thefts", style = "quantile", palette = "Blues", title = "Thefts") +
  tm_layout(legend.position = c("left", "bottom"))

#plot thefts per person by ward
tm_shape(toronto_ward_info) +
  tm_polygons(col = "Thefts_per_person", style = "cont", palette = "Blues", title = "Thefts per Person") +
  tm_layout(legend.position = c("left", "bottom"))

#plot thefts per person by ward
tm_shape(toronto_ward_info) +
  tm_polygons(col = "Thefts_per_person", style = "quantile", palette = "Blues", title = "Thefts per Person") +
  tm_layout(legend.position = c("left", "bottom"))

#plot thefts per sq km by ward
tm_shape(toronto_ward_info) +
  tm_polygons(col = "Thefts_per_sq_km", style = "cont", palette = "Blues", title = "Thefts per sq km") +
  tm_layout(legend.position = c("left", "bottom"))

#plot thefts per sq km by ward
tm_shape(toronto_ward_info) +
  tm_polygons(col = "Thefts_per_sq_km", style = "quantile", palette = "Blues", title = "Thefts per sq km") +
  tm_layout(legend.position = c("left", "bottom"))


#plot thefts per pop density by ward
tm_shape(toronto_ward_info) +
  tm_polygons(col = "Thefts_per_pop_density", style = "cont", palette = "Blues", title = "Thefts per Population Density") +
  tm_layout(legend.position = c("left", "bottom"))

#plot thefts per pop density by ward
tm_shape(toronto_ward_info) +
  tm_polygons(col = "Thefts_per_pop_density", style = "quantile", palette = "Blues", title = "Thefts per Population Density") +
  tm_layout(legend.position = c("left", "bottom"))



```


```{r, echo=FALSE, results='hide', message=FALSE, warning=FALSE}
toronto_basemap = tmaptools::read_osm(toronto_border, type='osm', mergeTiles=TRUE) 
```


```{r, echo = FALSE, message=FALSE, warning=FALSE, fig.cap = "Background raster displaying Toronto's subway stops."}
# Plot Toronto basemap with subway stops, each colour relates to it's line, plot black line for each subway line too
tm_shape(toronto_basemap) +
  tm_rgb() +
  tm_shape(toronto_subwayStops) +
  tm_dots(col = "subwayLine", size = 0.2, legend.show = FALSE, palette = "Set1") +
  tm_shape(toronto_subwayLines) +
  tm_lines(col = "name", size = 0.5, legend.show = FALSE, palette = "Set1")
```

```{r}
#plot toronto bus stops
tm_shape(toronto_basemap) +
  tm_rgb() +
  tm_shape(toronto_busStops) +
  tm_dots(col = "name", size = 0.2, legend.show = FALSE, palette = "Set1")
```

```{r}
#plot toronto bike parking
tm_shape(toronto_basemap) +
  tm_rgb() +
  tm_shape(toronto_bikeParking) +
  tm_dots(size = 0.2, legend.show = FALSE, palette = "Set1")

#plot toronto bike parking with thefts per pop density by ward under
tm_shape(toronto_ward_info) +
  tm_polygons(col = "Thefts_per_pop_density", style = "quantile", palette = "Blues", title = "Thefts per Population Density") +
  tm_shape(toronto_bikeParking) +
  tm_dots(size = 0.2, legend.show = FALSE, palette = "Set2", alpha = 0.25) +
  tm_layout(legend.position = c("left", "bottom"))
```

Analyzing bike crime in Toronto using your data can provide valuable insights into spatial and temporal patterns. Here are some methods you can consider:

### 1. **Spatial Analysis:**
   - **Hotspot Analysis:**
     - Use **Kernel Density Estimation (KDE)** to identify areas with high concentrations of bicycle thefts. This can help pinpoint hotspots where bike theft is more frequent.
     - Perform **Getis-Ord Gi* statistic** to detect statistically significant clusters of high or low bike theft occurrences.

   - **Proximity Analysis:**
     - Analyze the proximity of bike theft incidents to subway stops, bus stops, and bike parking facilities. This can reveal whether thefts are more likely to occur near certain types of infrastructure.
     - Use **buffer analysis** to create zones around bike parking facilities or transit stops and analyze theft rates within these zones.

   - **Correlation with Neighborhood Crime Rates:**
     - Investigate the correlation between bicycle thefts and overall neighborhood crime rates. This could help determine if bike thefts are more common in neighborhoods with higher crime rates.
     - Use **spatial regression models** (e.g., Geographically Weighted Regression, Spatial Lag Models) to explore how different spatial factors (e.g., proximity to transit, neighborhood crime rates) influence the likelihood of bicycle theft.

   - **Cluster Analysis:**
     - Perform **spatial clustering** (e.g., K-means, DBSCAN) to identify clusters of thefts and explore what characteristics these areas share, such as types of transit nearby or bike parking availability.

### 2. **Temporal Analysis:**
   - **Trend Analysis:**
     - Analyze the time series of bike theft incidents to identify trends over time, such as increases or decreases in thefts during specific months or years.
     - Decompose time series data to examine seasonal patterns, trends, and residuals using methods like **Seasonal Decomposition of Time Series (STL)**.

   - **Seasonal and Temporal Patterns:**
     - Examine seasonal trends in bike thefts, identifying whether certain times of the year (e.g., summer) see higher rates of theft.
     - Use **time-of-day analysis** to explore when thefts are most likely to occur, perhaps in relation to commuting patterns or daylight hours.

   - **Temporal Clustering:**
     - Perform **temporal clustering** to identify periods with spikes in bike thefts and analyze what factors might contribute to these spikes, such as events, holidays, or weather conditions.

### 3. **Spatio-Temporal Analysis:**
   - **Spatio-Temporal Hotspot Analysis:**
     - Extend hotspot analysis to the spatio-temporal domain by identifying not only where but also when thefts are most likely to occur. This can be done using **spatio-temporal KDE** or **Space-Time Cube analysis**.
     
   - **Spatio-Temporal Clustering:**
     - Implement spatio-temporal clustering methods (e.g., **ST-DBSCAN**) to identify patterns that emerge over space and time, such as recurring theft hotspots at particular times of the day or year.
     
   - **Spatio-Temporal Regression Models:**
     - Build spatio-temporal regression models to understand how spatial and temporal variables together influence bike thefts. This might include variables like weather, events, or changes in public transit schedules.

### 4. **Accessibility and Equity Analysis:**
   - **Accessibility Analysis:**
     - Analyze the accessibility of bike parking and its relation to thefts, considering whether areas with fewer parking facilities have higher theft rates.
     
   - **Equity Analysis:**
     - Investigate whether bike thefts are more prevalent in certain neighborhoods, possibly relating this to socio-economic factors or public infrastructure distribution.

### 5. **Predictive Modeling:**
   - **Machine Learning Models:**
     - Train machine learning models (e.g., Random Forest, Gradient Boosting) using spatial and temporal features to predict future bike theft occurrences. This can be useful for proactive policing and resource allocation.
     
   - **Risk Mapping:**
     - Develop a risk map indicating areas with a higher predicted risk of bike theft, combining spatial, temporal, and environmental factors.

By combining these methods, you can gain a comprehensive understanding of bike theft patterns in Toronto, leading to more informed decisions and effective interventions.


#Hotspot Analysis
#Kernel Density Estimation (KDE)

```{r}
#KDE for bicycle thefts
bicycle_thefts_sf = st_as_sf(BicycleThefts, coords = c("LONG_WGS84", "LAT_WGS84"), crs = 4326)
bicycle_thefts_transformed <- st_transform(bicycle_thefts_sf, crs = 32617)

toronto_border_st <- st_transform(toronto_border, crs = 32617)
window <- as.owin(toronto_border_st)

#convert to .ppp object
bicycle_thefts_ppp = bicycle_thefts_transformed |> 
  st_coordinates() |>
  as.ppp( W = window )

```

```{r}
kde <- density.ppp(bicycle_thefts_ppp) |>
  terra::rast()

crs(kde) = "EPSG:4547"
```

```{r}
#plot kde
tm_shape(kde) +
  tm_raster() +
  tm_shape(toronto_border) +
  tm_borders()
```

- **Proximity Analysis:**
     - Analyze the proximity of bike theft incidents to subway stops, bus stops, and bike parking facilities. This can reveal whether thefts are more likely to occur near certain types of infrastructure.
     - Use **buffer analysis** to create zones around bike parking facilities or transit stops and analyze theft rates within these zones.

```{r}
#Buffer analysis for bike parking

# Assuming you have already transformed bike thefts to UTM zone 17N (EPSG:32617)
# Transform other datasets to the same CRS
subway_stops <- st_transform(toronto_subwayStops, crs = 32617)
bus_stops <- st_transform(toronto_busStops, crs = 32617)
bike_parking <- st_transform(toronto_bikeParking, crs = 32617)
```

```{r}
# Add unique IDs to buffers
bike_parking_buffers <- st_buffer(bike_parking, dist = 200)
bike_parking_buffers$id <- seq_len(nrow(bike_parking_buffers))  # Unique ID for each buffer

subway_buffers <- st_buffer(subway_stops, dist = 200)
subway_buffers$id <- seq_len(nrow(subway_buffers))

bus_buffers <- st_buffer(bus_stops, dist = 200)
bus_buffers$id <- seq_len(nrow(bus_buffers))

```

```{r}
# Join bike thefts with bike parking buffers
thefts_in_parking_buffers <- st_join(bicycle_thefts_transformed, bike_parking_buffers)

# Summarize thefts within bike parking buffers
thefts_in_parking_summary <- thefts_in_parking_buffers %>%
  filter(!is.na(id)) %>%
  group_by(id) %>%
  summarize(num_thefts = n(), .groups = 'drop')

# Repeat for subway and bus buffers
thefts_in_subway_buffers <- st_join(bicycle_thefts_transformed, subway_buffers)
thefts_in_subway_summary <- thefts_in_subway_buffers %>%
  filter(!is.na(id)) %>%
  group_by(id) %>%
  summarize(num_thefts = n(), .groups = 'drop')

thefts_in_bus_buffers <- st_join(bicycle_thefts_transformed, bus_buffers)
thefts_in_bus_summary <- thefts_in_bus_buffers %>%
  filter(!is.na(id)) %>%
  group_by(id) %>%
  summarize(num_thefts = n(), .groups = 'drop')

```

```{r}
print(thefts_in_parking_summary)
print(thefts_in_subway_summary)
print(thefts_in_bus_summary)

```

```{r}
# Example for bike parking
ggplot() +
  geom_sf(data = bike_parking_buffers, fill = "black", alpha = 1) +
  geom_sf(data = thefts_in_parking_summary, color = "red", alpha = 0.1) +
  theme_minimal() +
  labs(title = "Bike Thefts near Bike Parking Facilities")

# Repeat for subway and bus stops
ggplot() +
  geom_sf(data = subway_buffers, fill = "black", alpha = 1) +
  geom_sf(data = thefts_in_subway_summary, color = "red", alpha = 0.1) +
  theme_minimal() +
  labs(title = "Bike Thefts near Subway Stops")

ggplot() +
  geom_sf(data = bus_buffers, fill = "black", alpha = 1) +
  geom_sf(data = thefts_in_bus_summary, color = "red", alpha = 0.1) +
  theme_minimal() +
  labs(title = "Bike Thefts near Bus Stops")

#plot bike thefts near bike parking facilities with population density under
tm_shape(toronto_ward_info) +
  tm_polygons(col = "Population_Density", style = "quantile", palette = "Blues", title = "Population Density (people/sq km)") +
  tm_shape(bike_parking_buffers) +
  tm_fill(col = "black", alpha = 0.5) +
  tm_shape(thefts_in_parking_summary) +
  tm_dots(col = "red", size = 0.1, alpha = 0.1) +
  tm_layout(legend.position = c("left", "bottom"))

```


```{r}
# Bar plot of theft counts in bike parking buffers
ggplot(thefts_in_parking_summary, aes(x = id, y = num_thefts)) +
  geom_bar(stat = "identity") +
  theme_minimal() +
  labs(title = "Thefts in Bike Parking Buffers", x = "Buffer ID", y = "Number of Thefts")

# Bar plot of theft counts in subway buffers
ggplot(thefts_in_subway_summary, aes(x = id, y = num_thefts)) +
  geom_bar(stat = "identity") +
  theme_minimal() +
  labs(title = "Thefts in Transit Buffers", x = "Buffer ID", y = "Number of Thefts")

# Bar plot of theft counts in bus buffers
ggplot(thefts_in_bus_summary, aes(x = id, y = num_thefts)) +
  geom_bar(stat = "identity") +
  theme_minimal() +
  labs(title = "Thefts in Bus Buffers", x = "Buffer ID", y = "Number of Thefts")

```

```{r}
# Combine summaries for bike parking and subway
combined_summary <- thefts_in_parking_summary %>%
  mutate(buffer_type = "Bike Parking") %>%
  bind_rows(
    thefts_in_subway_summary %>%
      mutate(buffer_type = "Subway")
  )

# Statistical test or regression model
# Example: t-test (assuming normality)
t.test(num_thefts ~ buffer_type, data = combined_summary)

# Repeat for bus stops
combined_summary <- thefts_in_parking_summary %>%
  mutate(buffer_type = "Bike Parking") %>%
  bind_rows(
    thefts_in_bus_summary %>%
      mutate(buffer_type = "Bus")
  )
    
t.test(num_thefts ~ buffer_type, data = combined_summary)

# Repeat for subway and bus stops
combined_summary <- thefts_in_subway_summary %>%
  mutate(buffer_type = "Subway") %>%
  bind_rows(
    thefts_in_bus_summary %>%
      mutate(buffer_type = "Bus")
  )
    
t.test(num_thefts ~ buffer_type, data = combined_summary)



```

### 2. **Temporal Analysis:**
   - **Trend Analysis:**
     - Analyze the time series of bike theft incidents to identify trends over time, such as increases or decreases in thefts during specific months or years.
     - Decompose time series data to examine seasonal patterns, trends, and residuals using methods like **Seasonal Decomposition of Time Series (STL)**.

   - **Seasonal and Temporal Patterns:**
     - Examine seasonal trends in bike thefts, identifying whether certain times of the year (e.g., summer) see higher rates of theft.
     - Use **time-of-day analysis** to explore when thefts are most likely to occur, perhaps in relation to commuting patterns or daylight hours.

   - **Temporal Clustering:**
     - Perform **temporal clustering** to identify periods with spikes in bike thefts and analyze what factors might contribute to these spikes, such as events, holidays, or weather conditions.

```{r}
# barplot of thefts by year
BicycleThefts %>%
  mutate(OCC_YEAR = as.numeric(as.character(OCC_YEAR))) %>%
  ggplot(aes(x = OCC_YEAR)) +
  geom_bar() +
  labs(title = "Bicycle Thefts by Year", x = "Year", y = "Number of Thefts")

#barplot of thefts by month
BicycleThefts %>%
  mutate(OCC_MONTH = month(OCC_DATE, label = TRUE)) %>%
  ggplot(aes(x = OCC_MONTH)) +
  geom_bar() +
  labs(title = "Bicycle Thefts by Month", x = "Month", y = "Number of Thefts")

#line graph of thefts over each for for entire period
ggplot(BicycleTheftsPerDay) +
  geom_line(aes(x = OCC_DATE_FROM2014, y = thefts)) +
  labs(title = "Bicycle Thefts Over Time", x = "Days from Start of 2014", y = "Number of Thefts")

# Plot the number of thefts per month over time
ggplot(BicycleTheftsPerMonth, aes(x = YearMonth, y = thefts, group = 1)) +
  geom_line(color = "blue") +
  geom_point() +  # Optional: Add points to highlight each month's data
  labs(title = "Monthly Bicycle Thefts Over Time",
       x = "Year",
       y = "Number of Thefts") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))  # Rotate x-axis labels for better readability

```

```{r}
#create predicting data for next 4 years
predict_data = cbind.data.frame(YearMonth = seq.Date(from = as.Date("2024-01-01"), to = as.Date("2027-12-01"), by = "month"),
                                time_id = seq(121, 168, 1),
                                time_id2 = seq(121, 168, 1),
                                thefts = rep(NA,48),
                                OCC_MONTH = month(seq.Date(from = as.Date("2024-01-01"), to = as.Date("2027-12-01"), by = "month")),
                                OCC_YEAR = year(seq.Date(from = as.Date("2024-01-01"), to = as.Date("2027-12-01"), by = "month"))
                                )

time_data_wmiss=rbind.data.frame(BicycleTheftsPerMonth, predict_data)
```

```{r}
time_model_seasonal = inla(thefts ~ 1 + f(time_id, model = "ar1") + f(time_id2, model = "seasonal", season.length = 12), 
                           data = time_data_wmiss, 
                           family = "gaussian", 
                           E = list(time_data_wmiss$time_id2),
                           control.predictor = list(compute=TRUE))

```

```{r}
time_data_wmiss$forcasts = time_model_seasonal$summary.fitted.values$mean
time_data_wmiss$lb = time_model_seasonal$summary.fitted.values[,3]
time_data_wmiss$ub = time_model_seasonal$summary.fitted.values[,5]
time_data_wmiss <- time_data_wmiss %>% ungroup()
```

```{r}
dataForcast <- time_data_wmiss |> filter (YearMonth > as.Date("2023-12-01"))
dataTheft <- time_data_wmiss |> filter (YearMonth <= as.Date("2023-12-01"))
```

```{r}

ggplot() +
  geom_line(data = dataForcast, aes(x = YearMonth, y = forcasts), col = "red", lwd = 0.8) +
  geom_ribbon(data = dataForcast, aes(ymin = lb, ymax = ub, x = YearMonth), fill = "red", alpha = 0.4) +
  geom_line(data = dataTheft, aes(x = YearMonth, y = thefts), col = "blue", lwd = 0.8) +
  theme_bw() +
  labs(y = "Number of Thefts", x = "Year")
```

```{r}
#KDE with population density
bicycle_thefts_with_wards_sf = st_as_sf(BicycleThefts_with_wards, coords = c("LONG_WGS84", "LAT_WGS84"), crs = 4326)
bicycle_thefts_with_wards_transformed <- st_transform(bicycle_thefts_with_wards_sf, crs = 32617)

toronto_border_st <- st_transform(toronto_border, crs = 32617)
window <- as.owin(toronto_border_st)

# Filter out rows with NA coordinates and outside points
valid_points <- st_filter(bicycle_thefts_with_wards_transformed, toronto_border_st)

# Ensure that points have valid coordinates
valid_points <- valid_points[!is.na(st_coordinates(valid_points)[,1]) & !is.na(st_coordinates(valid_points)[,2]), ]

# Convert valid points to ppp object
bicycle_thefts_with_wards_ppp <- as.ppp(
  st_coordinates(valid_points),
  W = window
)

# Attach the corresponding population density as weights
bicycle_thefts_with_wards_ppp$marks <- valid_points$Population_Density

# Perform KDE with Population Density as weights
kde_with_pop <- density.ppp(bicycle_thefts_with_wards_ppp, weights = bicycle_thefts_with_wards_ppp$marks) |>
  terra::rast()

# Ensure the CRS is correctly set
crs(kde_with_pop) <- "EPSG:32617"

# Plotting the KDE with population density
tm_shape(kde_with_pop) +
  tm_raster(palette = "YlOrRd") +
  tm_shape(toronto_border_st) +
  tm_borders()

```
### 5. **Predictive Modeling:**
   - **Machine Learning Models:**
     - Train machine learning models (e.g., Random Forest, Gradient Boosting) using spatial and temporal features to predict future bike theft occurrences. This can be useful for proactive policing and resource allocation.
     
   - **Risk Mapping:**
     - Develop a risk map indicating areas with a higher predicted risk of bike theft, combining spatial, temporal, and environmental factors.

By combining these methods, you can gain a comprehensive understanding of bike theft patterns in Toronto, leading to more informed decisions and effective interventions.

```{r}
#MLM

#regrab data so I remember what the hell I have
print(BicycleThefts_with_wards)
print(toronto_busStops)
print(toronto_subwayStops)
print(toronto_bikeParking)
print(toronto_subwayLines)
```


```{r}
#get distance from nearest bike parking

# Calculate distances from each bike theft to all bike parking facilities
distances_to_parking <- st_distance(BicycleThefts_with_wards, toronto_bikeParking)

# Convert distances to a matrix
distances_matrix <- as.matrix(distances_to_parking)

# Find the minimum distance for each bike theft (distance to the nearest bike parking)
BicycleThefts_with_wards$dist_to_bike_parking <- apply(distances_matrix, 1, min) |> units::set_units(m)

```

```{r}
#get distance from nearest subway stop and the subway line it is on (toronto_subwayStops$subwayLine)

# Calculate distances from each bike theft to all subway stops
distances_to_subway <- st_distance(BicycleThefts_with_wards, toronto_subwayStops)

# Convert distances to a matrix
distances_matrix <- as.matrix(distances_to_subway)

# Find the minimum distance for each bike theft (distance to the nearest subway stop)
BicycleThefts_with_wards$dist_to_subway <- apply(distances_matrix, 1, min) |> units::set_units(m)

# Find the subway line for each bike theft
BicycleThefts_with_wards$subway_line <- apply(distances_matrix, 1, function(x) toronto_subwayStops$subwayLine[which.min(x)])


```

```{r}
#get distance from nearest bus stop

# Calculate distances from each bike theft to all bus stops
distances_to_bus <- st_distance(BicycleThefts_with_wards, toronto_busStops)

# Convert distances to a matrix
distances_matrix <- as.matrix(distances_to_bus)

# Find the minimum distance for each bike theft (distance to the nearest bus stop)
BicycleThefts_with_wards$dist_to_bus <- apply(distances_matrix, 1, min) |> units::set_units(m)

```



```{r}
bike_theft_model_data <- BicycleThefts_with_wards |>
  dplyr::select(BIKE_TYPE, PREMISES_TYPE, OCC_DAY, OCC_YEAR, OCC_MONTH, OCC_HOUR, OCC_DOW, ward_id)#, geometry)

#if bike theft is in ward 13, 10, 11, or 14, set it to 1, else 0
bike_theft_model_data$is_downtown <- ifelse(bike_theft_model_data$ward_id %in% c("13", "10", "11", "14"), 1, 0)
bike_theft_model_data$is_downtown <- sapply(bike_theft_model_data$is_downtown, as.factor)

#remove ward_id
bike_theft_model_data <- bike_theft_model_data |>
  dplyr::select(-c(ward_id))

bike_theft_model_data$BIKE_TYPE <- sapply(bike_theft_model_data$BIKE_TYPE, as.factor)
bike_theft_model_data$PREMISES_TYPE <- sapply(bike_theft_model_data$PREMISES_TYPE, as.factor)
bike_theft_model_data$OCC_DAY <- sapply(bike_theft_model_data$OCC_DAY, as.factor)
bike_theft_model_data$OCC_YEAR <- sapply(bike_theft_model_data$OCC_YEAR, as.factor)
bike_theft_model_data$OCC_MONTH <- sapply(bike_theft_model_data$OCC_MONTH, as.factor)
bike_theft_model_data$OCC_HOUR <- sapply(bike_theft_model_data$OCC_HOUR, as.factor)
bike_theft_model_data$OCC_DOW <- sapply(bike_theft_model_data$OCC_DOW, as.factor)

# Add distance to subway, bus stop, and bike parking
bike_theft_model_data$dist_to_subway <- sapply(BicycleThefts_with_wards$dist_to_subway, as.numeric)
bike_theft_model_data$subway_line <- sapply(BicycleThefts_with_wards$subway_line, as.factor)
bike_theft_model_data$dist_to_bike_parking <- sapply(BicycleThefts_with_wards$dist_to_bike_parking, as.numeric)
bike_theft_model_data$dist_to_bus <- sapply(BicycleThefts_with_wards$dist_to_bus, as.numeric)
```

```{r}
# Convert sf object to data frame
bike_theft_model_data_df <- as.data.frame(bike_theft_model_data)

# Remove the geometry column
bike_theft_model_data_df <- bike_theft_model_data_df %>%
  dplyr::select(-geometry)

# If needed, convert it back to a standard data frame
bike_theft_model_data_clean <- as.data.frame(bike_theft_model_data_df)

#remove subways line column
bike_theft_model_data_clean <- bike_theft_model_data_clean |>
  dplyr::select(-c(subway_line))

#remove na values in ward_id
#bike_theft_model_data_clean <- bike_theft_model_data_clean |>
#  dplyr::filter(!is.na(ward_id))
```

```{r}
set.seed(123)  # For reproducibility
trainIndex <- createDataPartition(bike_theft_model_data_clean$is_downtown, p = 0.7, list = FALSE)
trainData <- bike_theft_model_data_clean[trainIndex, ]
testData <- bike_theft_model_data_clean[-trainIndex, ]
```

```{r}

# Train Random Forest model
rf_model <- randomForest(is_downtown ~ ., data = trainData, importance = TRUE)
 
predictions <- predict(rf_model, testData, type = "response")

print(confusionMatrix(predictions, testData$is_downtown))
print(importance(rf_model))
```

```{r}

# Prepare data for xgboost
dtrain <- xgb.DMatrix(data = as.matrix(trainData[, -which(names(trainData) == "TARGET_VARIABLE")]), label = trainData$TARGET_VARIABLE)
dtest <- xgb.DMatrix(data = as.matrix(testData[, -which(names(testData) == "TARGET_VARIABLE")]), label = testData$TARGET_VARIABLE)

params <- list(
  objective = "reg:squarederror",
  eval_metric = "rmse",
  eta = 0.1,
  max_depth = 6
)

xgb_model <- xgb.train(params, dtrain, nrounds = 100)
```

```{r}
# Example of visualizing predictions (this assumes you have future_data)
ggplot(data = future_data, aes(x = longitude, y = latitude, color = future_predictions_rf)) +
  geom_point() +
  labs(title = "Predicted Bike Theft Risk", x = "Longitude", y = "Latitude") +
  scale_color_viridis_c() +
  theme_minimal()
```

