---
title: "EDA - DATA*6500 Course Project Proposal"
author: "Evan Switzer, Namrata Roy"
date: "2024-07-11"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Introduction

A few years ago I(Evan) had my bike stolen at a Go Station. I now ride a beater bike that has got to weigh 50 pounds. I curse my old self every time I ride my "new" bike.

We aim to conduct an analysis of crime patterns in Toronto with a specific focus on subway stations and bike thefts. Although this is a narrow scope, we have expanded our project to explore major crime indicators such as  Assault, Auto Theft, Break and Enter, Robbery, Theft Over, Homicide, Shootings, etc. Our goal is to identify trends in various types of crime across Toronto neighborhoods and extract valuable insights from the data based on location.

To satisfy our purpose we are considering two more datasets along with the Bicycle Thefts dataset from the open source data of Toronto Police Service. One is the Neighborhood Crime Rates dataset and another is the Major Crime Indicators (MCI) dataset.
```{r, echo=FALSE, results='hide', message=FALSE, warning=FALSE}
install.packages("sf")
install.packages("dplyr")
install.packages("tmap")
install.packages("osmdata")
install.packages("rosm")
install.packages("tmaptools")
install.packages("ggplot2")
install.packages("tinytex")
install.packages("tidyr")
install.packages("leaflet")
```

```{r, echo=FALSE, results='hide', message=FALSE, warning=FALSE}
library(dplyr)
library(tmap)
library(osmdata)
library(sf)
library(rosm)
library(tmaptools)
library(ggplot2)
library(tinytex)
library(tidyr)
library(leaflet)
```

# Data Sources

## Neighbourhood_Crime_Rates_Open_Data
#### https://data.torontopolice.on.ca/datasets/TorontoPS::neighbourhood-crime-rates-open-data/about

The Toronto Neighbourhoods Boundary File contains detailed crime data categorized by neighbourhood. The data includes counts of offences and/or victims for Assault, Auto Theft, Break and Enter, Robbery, Theft Over, Homicide, Shootings, and Theft from Motor Vehicle. Additionally, the dataset features crime rates per 100,000 people for each neighbourhood, based on the projected population for each year as determined by Environics Analytics.

Our dataset on Neighborhood Crime Rates provides us with information on the frequency of different types of crime in each neighbourhood, along with the corresponding rate per 100,000 people. We have selected assault, auto theft, and bike theft as the categories for our analysis.


```{r, echo=FALSE, results='hide', message=FALSE, warning=FALSE}
#input data
NeighbourhoodCrimeRates = read.csv("Neighbourhood_Crime_Rates_Open_Data.csv")

MajorCrimeIndicators = read.csv("Major_Crime_Indicators_Open_Data.csv")

BicycleThefts = read.csv("Bicycle_Thefts_Open_Data.csv")
```

```{r, echo=FALSE, results='hide', message=FALSE, warning=FALSE}
#NeighbourhoodCrimeRates columns
names(NeighbourhoodCrimeRates)
#areas
NeighbourhoodCrimeRates$AREA_NAME
```

```{r, echo=FALSE, results='hide', message=FALSE, warning=FALSE}
#get background map of Toronto using openstreetmap
toronto_bbox = c(left = -79.6392, 
                 bottom = 43.403221,
                 right = -79.115952,
                 top = 43.855457)

toronto_border = opq(toronto_bbox, timeout = 600) |> 
  add_osm_feature(key = "boundary", value = "administrative") |> 
  add_osm_feature(key = 'admin_level', value = '6') |> 
  osmdata_sf() |> 
  (\(x) x$osm_multipolygons)() |>
  filter(name == "Toronto") |> 
  select()

toronto_ward = opq(toronto_bbox, timeout = 60) |> 
  add_osm_feature(key = 'admin_level', value = '9') |> 
  osmdata_sf() |> 
  (\(x) x$osm_multipolygons)()

subwayStops = opq(toronto_bbox) |> 
  add_osm_feature(key = 'railway', value = 'subway_entrance') |> 
  osmdata_sf() |> 
  (\(x) x$osm_points)() |> 
  select(name)

toronto_subwayStops = subwayStops[toronto_border,, op=st_within] 
```


```{r, echo=FALSE, results='hide', message=FALSE, warning=FALSE}
toronto_basemap = tmaptools::read_osm(toronto_border, type='osm', mergeTiles=TRUE) 
```

We load the base map of Toronto's Neighborhoods using the openstreetmap library. Figure 1 is the background raster of Toronto's Neighborhoods. Figure 2 shows the background raster of the neighborhood with all subway stops. 

```{r, echo = FALSE, fig.cap = "Background raster displaying Toronto's Neibourhoods"}
tm_shape(toronto_basemap)+
  tm_rgb()+
  tm_shape(toronto_ward)+
  tm_borders()
```

```{r, echo = FALSE, message=FALSE, warning=FALSE, fig.cap = "Background raster displaying Toronto's subway stops."}
# Plot Toronto basemap with subway stops
tm_shape(toronto_basemap) +
  tm_rgb() +
  tm_shape(toronto_subwayStops) +
  tm_dots(col = "name", size = 0.2, legend.show = FALSE)
```

```{r, echo=FALSE, results='hide', message=FALSE, warning=FALSE}
#MajorCrimeIndicators columns
names(MajorCrimeIndicators)

# Remove rows with "NSA" in any column
MajorCrimeIndicators = MajorCrimeIndicators |>
  filter_all(all_vars(. != "NSA"))
```

## Major_Crime_Indicators_Open_Data
#### https://data.torontopolice.on.ca/datasets/TorontoPS::major-crime-indicators-open-data/about

This dataset includes all Major Crime Indicators (MCI) occurrences by reported date and related offences since 2014.

The Major Crime Indicators (MCIs) encompass a range of categories, including Assault, Break and Enter, Auto Theft, Robbery, and Theft Over (excluding Sexual Violations). This data is provided at the offence and/or victim level, indicating that a single occurrence number may be linked to multiple rows of data associated with the various MCIs used to categorize the occurrence.

Our Major Crime Indicators dataset gives us a row for each offence committed since 2014. It contains all categories of major crime as well as location and date. In the figure 3 location of the major crimes and the categories of the crime are demonstrated.

```{r, echo=FALSE, fig.cap = "Major crime indicators and the categories of crime."}
# Convert MajorCrimeIndicators to an sf object using valid coordinates
MajorCrimeIndicators = st_as_sf(MajorCrimeIndicators, coords = c("LONG_WGS84", "LAT_WGS84"), crs = 4326)


# Ensure all spatial objects are in the same CRS
toronto_basemap = st_transform(toronto_basemap, crs = st_crs(MajorCrimeIndicators))
toronto_ward = st_transform(toronto_ward, crs = st_crs(MajorCrimeIndicators))

#plot MajorCrimeIndicators on toronto basemap
tm_shape(toronto_basemap) +
  tm_rgb() +
  tm_shape(toronto_ward) +
  tm_borders() +
  tm_shape(MajorCrimeIndicators) +
  tm_dots(col = "MCI_CATEGORY", size = 0.05, alpha = 0.3, palette = "Set1")
```

```{r, echo=FALSE, fig.cap = "Major crime indicators and the categories of crime for year 2014."}

# Filter data for a particular year, e.g., 2014
MajorCrimeIndicators_2014 = MajorCrimeIndicators |>
  filter(OCC_YEAR == 2014)


# Plot MajorCrimeIndicators for the year 2014 on Toronto basemap
tm_shape(toronto_basemap) +
  tm_rgb() +
  tm_shape(toronto_ward) +
  tm_borders() +
  tm_shape(MajorCrimeIndicators_2014) +
  tm_dots(col = "MCI_CATEGORY", size = 0.05, alpha = 0.3, palette = "Set1")
```

```{r, echo=FALSE, fig.cap = "Major crime indicators and the categories of crime for year 2024."}

# Filter data for a particular year, e.g., 2024
MajorCrimeIndicators_2024 = MajorCrimeIndicators |>
  filter(OCC_YEAR == 2024)


# Plot MajorCrimeIndicators for the year 2024 on Toronto basemap
tm_shape(toronto_basemap) +
  tm_rgb() +
  tm_shape(toronto_ward) +
  tm_borders() +
  tm_shape(MajorCrimeIndicators_2024) +
  tm_dots(col = "MCI_CATEGORY", size = 0.05, alpha = 0.3, palette = "Set1")
```

```{r, echo=FALSE, fig.cap = "Major crime rate changes over time."}
# Convert the data into a format suitable for plotting
crime_trends = MajorCrimeIndicators |>
  group_by(OCC_YEAR, MCI_CATEGORY) |>
  summarise(count = n(), .groups = 'drop')

# Convert the summarized data to a spatial data frame (sf)
crime_trends_sf = st_as_sf(crime_trends, coords = c("OCC_YEAR", "count"), crs = 4326)

# Plot the crime rate changes over time
ggplot(crime_trends, aes(x = OCC_YEAR, y = count, color = MCI_CATEGORY)) +
  geom_line() +
  labs(title = "Crime Rate Changes Over Time",
       x = "Year",
       y = "Number of Incidents",
       color = "Crime Category") +
  theme_minimal()
```

```{r, echo=FALSE, fig.cap = "Seasonal crime rate trends."}
# Extract the month from the OCC_DATE
MajorCrimeIndicators$OCC_MONTH = format(as.Date(MajorCrimeIndicators$OCC_DATE, format="%Y/%m/%d %H:%M:%S"), "%B")

# Summarize the crime data by month and category
crime_trends_month = MajorCrimeIndicators |>
  group_by(OCC_MONTH, MCI_CATEGORY) |>
  summarise(count = n(), .groups = 'drop')

# Order the months correctly
crime_trends_month$OCC_MONTH = factor(crime_trends_month$OCC_MONTH, levels = month.name)

# Plot the crime rate changes by month 
ggplot(crime_trends_month, aes(x = OCC_MONTH, y = count, color = MCI_CATEGORY)) +
  geom_line() +
  geom_point() +
  labs(title = "Seasonal Crime Rate Trends",
       x = "Month",
       y = "Number of Incidents",
       color = "Crime Category") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

```{r, echo=FALSE, fig.cap = "Major Crime Rate by Toronto Ward."}
# Perform spatial join to associate each crime with its respective ward
MajorCrimeIndicators_ward = st_join(MajorCrimeIndicators, toronto_ward, join = st_within)

# Summarize the crime data by ward
crime_trends_ward = MajorCrimeIndicators_ward |>
  group_by(osm_id) |>
  summarise(count = n(), .groups = 'drop')

# Merge the aggregated data with the spatial data
crime_trends_toronto_ward = st_join(toronto_ward, crime_trends_ward, join = st_intersects)

# Replace NA values with 0 in the count column
crime_trends_toronto_ward$count[is.na(crime_trends_toronto_ward$count)] = 0

# Plot the crime rate by ward
tm_shape(toronto_basemap) +
  tm_rgb() +
  tm_shape(crime_trends_toronto_ward) +
  tm_polygons("count", palette = "Reds", title = "Crime Count") +
  tm_layout(title = "Crime Rate by Toronto Ward",
            legend.outside = TRUE)
```

```{r, echo=FALSE, fig.cap = "Theft Over Crime Rate by Toronto Ward."}
# Specify the crime types to filter
crime_types = c("Theft Over")

# Filter the crime data by the specified crime types
filtered_crimes = MajorCrimeIndicators |>
  filter(MCI_CATEGORY %in% crime_types)

# Perform spatial join to associate each crime with its respective ward
filtered_crimes_ward = st_join(filtered_crimes, toronto_ward, join = st_within)

# Summarize the filtered crime data by ward using the appropriate column
filtered_crime_trends_ward = filtered_crimes_ward |>
  group_by(name) |>
  summarise(count = n(), .groups = 'drop')

# Merge the aggregated data with the spatial data
filtered_crime_trends_toronto_ward = st_join(toronto_ward, filtered_crime_trends_ward, join = st_intersects)

# Replace NA values with 0 in the count column
filtered_crime_trends_toronto_ward$count[is.na(filtered_crime_trends_toronto_ward$count)] = 0

# Plot the filtered crime rate by ward
tm_shape(toronto_basemap) +
  tm_rgb() +
  tm_shape(filtered_crime_trends_toronto_ward) +
  tm_polygons("count", palette = "Reds", title = "Crime Count") +
  tm_layout(title = "Filtered Crime Rate by Toronto Ward",
            legend.outside = TRUE)
```

```{r, echo=FALSE, results='hide', message=FALSE, warning=FALSE}
#BicycleThefts columns
names(BicycleThefts)
#print any rows that have LONG_WGS84 or LAT_WGS84 as NA or 0
BicycleThefts |>
  filter(LONG_WGS84 == 0 | is.na(LONG_WGS84) | LAT_WGS84 == 0 | is.na(LAT_WGS84))
#remove the rows with NA or 0 in LONG_WGS84 or LAT_WGS84
BicycleThefts = BicycleThefts |>
  filter(LONG_WGS84 != 0 & !is.na(LONG_WGS84) & LAT_WGS84 != 0 & !is.na(LAT_WGS84))
```

## Bicycle_Thefts_Open_Data
#### https://data.torontopolice.on.ca/datasets/TorontoPS::bicycle-thefts-open-data/about

The dataset comprises records of reported cases of bicycle thefts dating back to 2014. The occurrences encompass a wide range of criminal activities where the theft of a bicycle was involved.

The Bicycle Thefts dataset provides detailed information on each incident of bicycle theft since 2014, including the theft's location, date, make, model, cost, and other pertinent details. Figure 4 depicts the overview of this dataset according to the wards of Toronto.

```{r, echo=FALSE, fig.cap = "Bicycle thefts and what type of premises they were stolen from."}
# Convert BicycleThefts to an sf object using valid coordinates
BicycleThefts = st_as_sf(BicycleThefts, coords = c("LONG_WGS84", "LAT_WGS84"), crs = 4326)

#plot BicycleThefts on toronto basemap
tm_shape(toronto_basemap) +
  tm_rgb() +
  tm_shape(toronto_ward) +
  tm_borders() +
  tm_shape(BicycleThefts) +
  tm_dots(col = "PREMISES_TYPE", size = 0.05, alpha = 0.3, palette = "Set1")
```


```{r, echo=FALSE}
# Convert OCC_DATE to Date format and extract month and year
BicycleThefts$OCC_DATE = as.Date(BicycleThefts$OCC_DATE, format="%Y/%m/%d %H:%M:%S")
BicycleThefts$Year = format(BicycleThefts$OCC_DATE, "%Y")
BicycleThefts$Month = format(BicycleThefts$OCC_DATE, "%B")

# Order the months correctly
BicycleThefts$Month = factor(BicycleThefts$Month, levels = month.name)

# Summarize the data by year for bicycle thefts
bike_thefts_year = BicycleThefts |>
  group_by(Year) |>
  summarise(count = n(), .groups = 'drop')

# Summarize the data by month
monthly_bicycle_thefts_trends = BicycleThefts |>
  group_by(Month) |>
  summarise(count = n(), .groups = 'drop')
```

```{r, echo=FALSE, fig.cap = "Bicycle Thefts Trends Over the year"}
# Plot bike theft trends over the years
ggplot(bike_thefts_year, aes(x = Year, y = count)) +
  geom_line(group = 1, color = "blue") +
  geom_point(color = "blue") +
  labs(title = "Bike Theft Trends Over the Years",
       x = "Year",
       y = "Number of Bike Thefts") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```

```{r, echo=FALSE, fig.cap = "Seasonal Trends in Bicycle Thefts"}
# Plot seasonal trends in bicycle thefts
ggplot(monthly_bicycle_thefts_trends, aes(x = Month, y = count)) +
  geom_line(group = 1, color = "blue") +
  geom_point(color = "blue") +
  labs(title = "Seasonal Trends in Bicycle Thefts",
       x = "Month",
       y = "Number of Thefts") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

```{r}
# Transform to the same CRS as BicycleThefts
toronto_neighborhoods = st_transform(toronto_ward, crs = st_crs(BicycleThefts))

# Summarize the data by neighborhood for bicycle thefts
bicycle_thefts_neighborhood = BicycleThefts |>
  group_by(NEIGHBOURHOOD_140) |>
  summarise(count = n())

# Merge with spatial data
combined_neighborhoods = st_join(toronto_neighborhoods, bicycle_thefts_neighborhood, join = st_within)

# Replace NA values with 0 in the count column
combined_neighborhoods$count[is.na(combined_neighborhoods$count)] = 0

# Create the interactive map
leaflet(combined_neighborhoods) |>
  addTiles() |>
  addPolygons(fillColor = ~colorNumeric("Blues", count)(count),
              fillOpacity = 0.7,
              color = "white",
              weight = 1,
              popup = ~paste("Neighborhood:", NEIGHBOURHOOD_140, "<br>",
                             "Bike Thefts:", count)) |>
  addLegend(pal = colorNumeric("Blues", combined_neighborhoods$count),
            values = combined_neighborhoods$count,
            title = "Bike Thefts",
            position = "bottomright")

```

```{r, echo=FALSE, fig.cap = "Seasonal Trends in Bicycle Thefts vs Major Crimes."}
# Summarize the data by month for bicycle thefts
bicycle_trends = BicycleThefts |>
  group_by(Month) |>
  summarise(count = n()) |>
  mutate(Crime_Type = "Bicycle Theft")

major_crime_trends_renamed = MajorCrimeIndicators |>
  rename(Month = OCC_MONTH)

# Summarize the data by month for major crimes
major_crime_trends = major_crime_trends_renamed |>
  group_by(Month, MCI_CATEGORY) |>
  summarise(count = n())|>
  rename(Crime_Type = MCI_CATEGORY)

# Combine the two datasets
combine_crime_trends = bind_rows(bicycle_trends, major_crime_trends)

# Plot seasonal trends in bicycle thefts and major crimes
ggplot(combine_crime_trends, aes(x = Month, y = count, color = Crime_Type, group = Crime_Type)) +
  geom_line() +
  geom_point() +
  labs(title = "Seasonal Trends in Bicycle Thefts vs Major Crimes",
       x = "Month",
       y = "Number of Incidents") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

```{r, echo=FALSE, fig.cap = "Seasonal Trends in Bicycle Thefts vs Major Crimes."}
# Summarize the data by neighborhood for bicycle thefts
bicycle_trends_neighborhood = BicycleThefts |>
  group_by(NEIGHBOURHOOD_140) |>
  summarise(count = n()) |>
  mutate(Crime_Type = "Bicycle Theft")

# Summarize the data by neighborhood for major crimes
major_crime_trends_neighborhood = MajorCrimeIndicators |>
  group_by(NEIGHBOURHOOD_140, MCI_CATEGORY) |>
  summarise(count = n()) |>
  rename(Crime_Type = MCI_CATEGORY)

# Combine the two datasets
combine_crime_trends_neighborhood = bind_rows(bicycle_trends_neighborhood, major_crime_trends_neighborhood)

# Plot crime rates by neighborhood
ggplot(combine_crime_trends_neighborhood, aes(x = reorder(NEIGHBOURHOOD_140, -count), y = count, fill = Crime_Type)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(title = "Crime Rates by Neighborhood",
       x = "Neighborhood",
       y = "Number of Incidents") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))
```

```{r}
# Rename column in new objects
bicycle_trends_neighborhood_count = bicycle_trends_neighborhood |>
  rename(bike_thefts_count = count)

major_crime_trends_neighborhood_count = major_crime_trends_neighborhood |>
  rename(total_crimes_count = count)
 
# Combine the two datasets
combined_crime_trends_neighborhood = st_join(bicycle_trends_neighborhood_count, major_crime_trends_neighborhood_count, join = st_intersects)

# Calculate the ratio of bike thefts to overall crimes
combined_crime_trends_neighborhood = combined_crime_trends_neighborhood |>
  mutate(bike_theft_ratio = bike_thefts_count / total_crimes_count)

# Plot the ratio of bike thefts to overall crimes by neighborhood
ggplot(combined_crime_trends_neighborhood, aes(x = reorder(NEIGHBOURHOOD_140.x, -bike_theft_ratio), y = bike_theft_ratio)) +
  geom_bar(stat = "identity", fill = "steelblue") +
  labs(title = "Ratio of Bike Thefts to Overall Crimes by Neighborhood",
       x = "Neighborhood",
       y = "Bike Theft Ratio") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))
```

```{r}
# Merge with spatial data
combined_crime_trends_neighborhood_sf = st_join(toronto_ward, combined_crime_trends_neighborhood)

# Replace NA values with 0 in the ratio column
combined_crime_trends_neighborhood_sf$bike_theft_ratio[is.na(combined_crime_trends_neighborhood_sf$bike_theft_ratio)] = 0

# Plot the ratio of bike thefts to overall crimes by neighborhood using tmap
tm_shape(combined_crime_trends_neighborhood_sf) +
  tm_polygons("bike_theft_ratio", palette = "Blues", title = "Bike Theft Ratio") +
  tm_layout(title = "Ratio of Bike Thefts to Overall Crimes by Neighborhood",
            legend.outside = TRUE)
```

```{r}

```

# Data Cleaning and Preperation

The  data is mostly clean since it is provided by  the Toronto Police Service,  but there are still some issues that need attention. For instance, some entries have the value NSA, which stands for "Not Specified Area." These entries will be removed since they are not significant in number and lack location information. Additionally, any rows with NA or 0 in the longitude or latitude columns will be excluded.

Furthermore, we have ensured that all spatial data has been converted to the same CRS for unified plotting.

# Methods

We will implement several Spacial Machine Learning Models such as binomial regression, Adaboosting or Gaussian process, etc., resample the methods and compare the outcome to consider the best model to satisfy our objective. We will apply point pattern analysis to find the trends for location based crime pattern. Additionally, we will update our methods to achieve our goal.

```{r, echo=FALSE, results='hide', message=FALSE, warning=FALSE}
#bar plot of crime rate ASSAULT_2014 x AREA_NAME (top 10)
# NeighbourhoodCrimeRates |>
#   top_n(10, ASSAULT_2014) |>
#   ggplot(aes(x = reorder(AREA_NAME, ASSAULT_2014), y = ASSAULT_2014, fill = AREA_NAME)) +
#   geom_bar(stat = "identity") +
#   coord_flip() +
#   labs(title = "Crime Rate of Assault in 2014 by Area",
#        x = "Area Name",
#        y = "Crime Rate") +
#   theme_classic() +
#   scale_fill_viridis_d()
# 
# #bar plot of crime rate ASSAULT_2014 x AREA_NAME (bottom 10)
# NeighbourhoodCrimeRates |>
#   top_n(-10, ASSAULT_2014) |>
#   ggplot(aes(x = reorder(AREA_NAME, ASSAULT_2014), y = ASSAULT_2014, fill = AREA_NAME)) +
#   geom_bar(stat = "identity") +
#   coord_flip() +
#   labs(title = "Crime Rate of Assault in 2014 by Area",
#        x = "Area Name",
#        y = "Crime Rate") +
#   theme_classic() +
#   scale_fill_viridis_d()
# 
# 
# #bar plot of POPULATION_2023 x AREA_NAME (top 10)
# NeighbourhoodCrimeRates |>
#   top_n(10, POPULATION_2023) |>
#   ggplot(aes(x = reorder(AREA_NAME, POPULATION_2023), y = POPULATION_2023, fill = AREA_NAME)) +
#   geom_bar(stat = "identity") +
#   coord_flip() +
#   labs(title = "Population in 2023 by Area",
#        x = "Area Name",
#        y = "Population") +
#   theme_classic() +
#   scale_fill_viridis_d()
# 
# #bar plot of POPULATION_2023 x AREA_NAME (bottom 10)
# NeighbourhoodCrimeRates |>
#   top_n(-10, POPULATION_2023) |>
#   ggplot(aes(x = reorder(AREA_NAME, POPULATION_2023), y = POPULATION_2023, fill = AREA_NAME)) +
#   geom_bar(stat = "identity") +
#   coord_flip() +
#   labs(title = "Population in 2023 by Area",
#        x = "Area Name",
#        y = "Population") +
#   theme_classic() +
#   scale_fill_viridis_d()

```


